Temporal Dynamics of Normalization Reweighting

### 2023-07-31 manuscript.tex
Implemented reviewer suggestions

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{lineno}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Temporal Dynamics of Normalization Reweighting},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Suppressive interactions between neurons are ubiquitous in the nervous system, with normalization considered a canonical neuronal computation (Carandini and Heeger, 2011). Yet for decades the strength of suppression was treated as fixed, due to the observation that adapting to one stimulus does not decrease its suppressive potency (Foley and Chen, 1997; Freeman et al., 2002). This orthodoxy has been challenged by a series of innovative studies showing that normalization can be `reweighted' by recent history (Aschner et al., 2018; Westrick et al., 2016; Yiltiz et al., 2020). When pairs of stimuli are repeatedly presented together, their neural representations come to suppress each other more strongly. Far from being fixed, normalization is therefore a dynamic process that is continuously updated by the sensory environment. Our objectives were to measure the timecourse of these changes non-invasively in the human brain, compare them across distinct suppressive pathways, and determine whether they differ as a function of autistic traits.

Plastic changes within the visual system occur over multiple timescales (see Webster, 2015 for a recent review). Cortical forms of adaptation to cues such as stimulus contrast (Blakemore and Campbell, 1969), orientation (Gibson and Radner, 1937) and motion (Mather et al., 2008) can be observed within a few seconds, but also build up over durations on the order of minutes (Greenlee et al., 1991). Other types of adaptation have been identified where changes occur over longer time periods, such as several hours (Kwon et al., 2009) or days (Haak et al., 2014). Previous normalization reweighting studies involved adapting sequences of around 40 -- 60s (Aschner et al., 2018; Yiltiz et al., 2020), but in principle reweighting might occur faster than this, consistent with other types of contrast adaptation.

Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation (Foley, 1994; Heeger, 1992), eye-of-origin (Legge, 1979; Sengpiel and Blakemore, 1994) and spatial position (Cannon and Fullenkamp, 1991; Petrov, 2005). At present there is evidence of normalization reweighting between stimuli with orthogonal orientations (Aschner et al., 2018), and adjacent spatial positions (Yiltiz et al., 2020). We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways (Li et al., 2005; e.g. Meese and Baker, 2009; Sengpiel and Vorobyov, 2005). Comparing monocular and dichoptic suppression permits any contribution of early pre-cortical factors to be isolated. This is because interocular suppression impacts in primary visual cortex, and bypasses any retinal and subcortical stages of processing that contribute to monocular suppression (Li et al., 2005).

Atypical sensory experience, including hypersensitivity to loud sounds, bright lights and strong odours or flavours, is widely reported by individuals on the autism spectrum (MacLennan et al., 2022; Simmons et al., 2009), but the causal mechanisms remain unclear. Fundamental measures of sensitivity including visual acuity (Tavassoli et al., 2011), contrast sensitivity (Koh et al., 2010), and audiometric performance (Rosenhall et al., 1999) are not consistently different from neurotypical controls. Theoretical accounts of sensory differences in autism have proposed that the balance of inhibition and excitation may be disrupted (Rosenberg et al., 2015; Rubenstein and Merzenich, 2003), yet the evidence is currently inconclusive (Sandhu et al., 2020; e.g. Schallmo et al., 2020; Van de Cruys et al., 2018). Our recent work identified an autism-related difference using steady-state EEG (Vilidaite et al., 2018), in which nonlinear (second harmonic) responses were weaker, implicating atypical suppression in autism.

@@

Experiment 1 was completed by 100 adult participants (32 male, 68 female; mean age 21.9) in early 2015, and first reported by Vilidaite et al. (2018). Here we reanalysed the dataset, and report the results of masking conditions not previously published. Experiment 2 was completed by 100 adult participants (23 male, 74 female, 3 other/not stated; mean age 22.1) in early 2022. Experiment 3 was completed by 10 adults (2 male, 8 female) with a clinical diagnosis of autism, and 10 control participants who were closely matched for age (means of 21.8 and 22, \emph{t} = 0.18, \emph{df} = 18, \emph{p} = 0.86) and exactly matched for gender. Procedures in Experiments 1 and 2 were approved by the ethics committee of the Department of Psychology at the University of York. Procedures for Experiment 3 were approved by the ethics committee of the York Neuroimaging Centre. All participants provided written informed consent, and procedures were consistent with the Declaration of Helsinki.

@@

In Experiment 3, stimuli were presented using a ProPixx DLP projector (VPixx Technologies) running at 120Hz. Stereo presentation was enabled using a circular polariser that was synchronised with the projector refresh, and participants wore passive polarised glasses during the experiment. DLP projectors are perfectly linear, so gamma correction was not required. Data were acquired using a refurbished 248-channel 4D Neuroimaging Magnes 3600 MEG scanner, recording at 1001Hz. Participant head shape was digitised using a Polhemus Fastrak device, and head position was recorded at the start and end of each block by passing current through 5 position coils placed at fiducial points on the head. We also obtained structural MRI scans using a 3 Tesla Siemens Magnetom Prisma scanner to aid in source localisation. Two participants were not available for MRI scans, so we used the MNI ICBM152 template brain (Fonov et al., 2011) for these individuals.

@@

Participants also completed either the short AQ (Hoekstra et al., 2011) in Experiment 1, or the full AQ (Baron-Cohen et al., 2001) in Experiments 2 and 3. For comparison across experiments, we rescaled the short AQ to the same range as the full AQ (0-50). In Experiments 2 and 3, the sensory perception quotient (SPQ) questionnaire (Tavassoli et al., 2014) was also completed.

@@

EEG data from Experiments 1 and 2 were first imported into Matlab using components of the EEGlab toolbox (Delorme and Makeig, 2004), and converted into a compressed ASCII format. Primary data analysis was then conducted using a bespoke \emph{R} script. In brief, we epoched each trial and extracted the average timecourse across four occipital electrodes (\emph{Oz}, \emph{POz}, \emph{O1} and \emph{O2}), and then calculated the Fourier transform of this average waveform. We excluded trials for which the Mahalanobis distance of the complex Fourier components exceeded 3 (for details see Baker, 2021). This resulted in 0.25\% of trials being excluded for Experiment 1, and 4.51\% of trials for Experiment 2. Next we averaged the waveforms across all remaining trials, and calculated the Fourier transform in a 1-second sliding window to generate timecourses for each participant. We divided the timecourse for the target-only condition by the timecourse for the target + mask condition to produce a suppression ratio. These were then converted to logarithmic (dB) units for averaging, calculation of standard errors, and statistical comparisons. For display purposes we smoothed the timecourses using a cubic spline function, however all statistical comparisons used the unsmoothed data.

For Experiment 3, we performed source localisation using a linearly constrained minimum variance (LCMV) beamformer algorithm, implemented in Brainstorm (Tadel et al., 2011). Structural MRI scans were processed using Freesurfer (Dale et al., 1999) to generate a 3D mesh of the head and brain, and we calculated source weights for each block with reference to a 5-minute empty room recording, usually recorded on the day of the experiment. The matrix of source weights for each block was used in a custom Matlab script to extract signals from V1, identified using the probabilistic maps of Wang et al. (2015). These signals were then imported into R for the main analysis, which was consistent with the EEG analysis described above. The outlier rejection procedure excluded 2.47\% of trials for Experiment 3.

To make comparisons between groups of participants across time, we used a non-parametric cluster correction technique (Maris and Oostenveld, 2007) based on t-tests. Clusters were identified as temporally adjacent observations that were all statistically significant, and a summed t-value was calculated for each cluster. A null distribution was then generated by randomising group membership and recalculating the summed t-value for the largest cluster, and repeating this procedure 1000 times. Clusters were considered significant if they fell outside of the 95\% confidence limits of the null distribution. We adapted this approach to test for significantly increasing suppression by conducting one-way t-tests between time points separated by 1000ms, and repeating the cluster correction procedure as described above.

@@

We began by reanalysing data from a steady-state visually evoked potential (SSVEP) experiment reported by Vilidaite et al. (2018). Participants viewed arrays of flickering gratings of varying contrasts. In some conditions a single grating orientation was present flickering at 7Hz, whereas in other conditions a high contrast `mask' was added at right angles to the target gratings, and flickering at 5Hz. The left panel of Figure \ref{fig:Pilotdata}a shows contrast response functions with and without the mask - the presence of the mask reduces the 7Hz response to the target (blue squares are below the black circles; significant main effect of mask contrast, F(1,99) = 26.52, \emph{p} \textless{} 0.001). Similarly, the right panel of Figure \ref{fig:Pilotdata}a shows that the 5Hz response to the mask was itself suppressed by the presence of high contrast targets (main effect of target contrast on the mask response, F(2.92,288.63) = 46.77, \emph{p} \textless{} 0.001; note that the data from the masking conditions were not reported by Vilidaite et al. (2018)). At both frequencies, responses were localised to the occipital pole (see insets).

@@

{\centering \includegraphics{Figures/VilidaiteData}

@@

\caption{Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking the electrodes over which signals were averaged (Oz, POz, O1 and O2). Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to baseline (grey). Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Points around y = 1.8 indicate a significantly increasing ratio. Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE across N=100 participants, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate the conditions used for subsequent analyses.}\label{fig:Pilotdata}
\end{figure}

We then performed a timecourse analysis, in which we analysed each 11-second trial using a sliding 1-second time window. The top panel of Figure \ref{fig:Pilotdata}c shows the response at the target frequency (7Hz) to a single stimulus of 32\% contrast (black), and the response at 7Hz when the 32\% contrast mask is added (blue). For comparison, a baseline timecourse is also shown (grey), which was the response at 7Hz when a 5Hz stimulus was shown (therefore controlling for attention, blinking etc.). Analogous responses are shown at three other frequencies - the mask frequency (5Hz), and the second harmonics of both target and mask frequencies (14Hz, 10Hz), at which strong responses were also found (see spectra in Figure \ref{fig:Pilotdata}b). The reduction in signal strength when the mask component is added illustrates the masking effect.

Taking the ratio of the two timecourses to calculate a masking index reveals that at 7Hz masking increases steeply during the first two seconds of stimulus presentation, and then plateaus for several seconds (blue trace in Figure \ref{fig:Pilotdata}d). A similar pattern is observed at 5Hz (red trace in Figure \ref{fig:Pilotdata}d), as well as at the second harmonics, with some variability in the timecourse across frequencies; for example, at 5Hz suppression peaks at around 4 seconds. The black trace shows the average masking ratio across all four frequencies, which rises steeply for just over two seconds and then stays approximately constant until stimulus offset. We conducted cluster-corrected t-tests between ratios separated by 1000ms, testing for an increase in suppression ratio across time (i.e.~a one-sided test). Points at y = 1.8 in Figure \ref{fig:Pilotdata}d indicate time points where the ratio is significantly increasing, and occur mostly during the first 2 seconds of stimulus presentation.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{lineno}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Temporal Dynamics of Normalization Reweighting},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Suppressive interactions between neurons are ubiquitous in the nervous system, with normalization considered a canonical neuronal computation (\protect\hyperlink{ref-Carandini2011}{Carandini \& Heeger, 2011}). One consequence of normalization is that neurons tuned to different stimulus features modulate each others' firing, usually via a process of divisive suppression (\protect\hyperlink{ref-Heeger1992}{Heeger, 1992}). For decades the strength of suppression was treated as fixed, due to the observation that adapting to one stimulus does not decrease its suppressive potency (\protect\hyperlink{ref-Foley1997}{Foley \& Chen, 1997}; \protect\hyperlink{ref-Freeman2002}{Freeman et al., 2002}). This orthodoxy has been challenged by a series of innovative studies proposing that normalization can be `reweighted' by recent history (\protect\hyperlink{ref-Aschner2018}{Aschner et al., 2018}; \protect\hyperlink{ref-Westrick2016}{Westrick et al., 2016}; \protect\hyperlink{ref-Yiltiz2020}{Yiltiz et al., 2020}). When pairs of stimuli are repeatedly presented together, their neural representations appear to suppress each other more strongly. Far from being fixed, normalization may therefore be a dynamic process that is continuously updated by the sensory environment. Our objectives were to measure the timecourse of changes in suppression non-invasively in the human brain, compare them across distinct anatomical pathways, and determine whether they differ as a function of autistic traits.

Plastic changes within the visual system occur over multiple timescales (see \protect\hyperlink{ref-Webster2015}{Webster, 2015} for a recent review). Cortical forms of adaptation to cues such as stimulus contrast (\protect\hyperlink{ref-Blakemore1969}{Blakemore \& Campbell, 1969}), orientation (\protect\hyperlink{ref-Gibson1937}{Gibson \& Radner, 1937}) and motion (\protect\hyperlink{ref-Mather2008}{Mather et al., 2008}) can be observed within a few seconds, but also build up over durations on the order of minutes (\protect\hyperlink{ref-Greenlee1991}{Greenlee et al., 1991}). Other types of adaptation have been identified where changes occur over longer time periods, such as several hours (\protect\hyperlink{ref-Kwon2009}{Kwon et al., 2009}) or days (\protect\hyperlink{ref-Haak2014}{Haak et al., 2014}). Previous normalization reweighting studies involved adapting sequences of around 40 -- 60s (\protect\hyperlink{ref-Aschner2018}{Aschner et al., 2018}; \protect\hyperlink{ref-Yiltiz2020}{Yiltiz et al., 2020}), but in principle reweighting might occur faster than this, consistent with other types of contrast adaptation.

Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation (\protect\hyperlink{ref-Foley1994}{Foley, 1994}; \protect\hyperlink{ref-Heeger1992}{Heeger, 1992}), eye-of-origin (\protect\hyperlink{ref-Dougherty2019}{Dougherty et al., 2019}; \protect\hyperlink{ref-Legge1979}{Legge, 1979}; \protect\hyperlink{ref-Sengpiel1994}{Sengpiel \& Blakemore, 1994}) and spatial position (\protect\hyperlink{ref-Cannon1991}{Cannon \& Fullenkamp, 1991}; \protect\hyperlink{ref-Petrov2005}{Petrov, 2005}). At present there is evidence of normalization reweighting between stimuli with orthogonal orientations (\protect\hyperlink{ref-Aschner2018}{Aschner et al., 2018}), and adjacent spatial positions (\protect\hyperlink{ref-Yiltiz2020}{Yiltiz et al., 2020}). We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways (\protect\hyperlink{ref-Li2005}{Li et al., 2005}; e.g. \protect\hyperlink{ref-Meese2009}{Meese \& Baker, 2009}; \protect\hyperlink{ref-Sengpiel2005}{Sengpiel \& Vorobyov, 2005}). Comparing monocular and dichoptic suppression permits any contribution of early pre-cortical factors to be isolated. This is because interocular suppression is generally thought to impact in primary visual cortex (though see \protect\hyperlink{ref-Dougherty2019}{Dougherty et al., 2019}), and bypasses any retinal and subcortical stages of processing that contribute to monocular suppression (\protect\hyperlink{ref-Li2005}{Li et al., 2005}).

Atypical sensory experience, including hypersensitivity to loud sounds, bright lights and strong odours or flavours, is widely reported by individuals on the autism spectrum (\protect\hyperlink{ref-MacLennan2022}{MacLennan et al., 2022}; \protect\hyperlink{ref-Simmons2009}{Simmons et al., 2009}), but the causal mechanisms remain unclear. Fundamental measures of sensitivity including visual acuity (\protect\hyperlink{ref-Tavassoli2011}{Tavassoli et al., 2011}), contrast sensitivity (\protect\hyperlink{ref-Koh2010}{Koh et al., 2010}), and audiometric performance (\protect\hyperlink{ref-Rosenhall1999}{Rosenhall et al., 1999}) are not consistently different from neurotypical controls. Theoretical accounts of sensory differences in autism have proposed that the balance of inhibition and excitation may be disrupted (\protect\hyperlink{ref-Rosenberg2015}{Rosenberg et al., 2015}; \protect\hyperlink{ref-Rubenstein2003}{Rubenstein \& Merzenich, 2003}), yet the evidence is currently inconclusive (\protect\hyperlink{ref-Sandhu2020}{Sandhu et al., 2020}; e.g. \protect\hyperlink{ref-Schallmo2020}{Schallmo et al., 2020}; \protect\hyperlink{ref-VandeCruys2018}{Van de Cruys et al., 2018}). Our recent work identified an autism-related difference using steady-state EEG (\protect\hyperlink{ref-Vilidaite2018}{Vilidaite et al., 2018}), in which nonlinear (second harmonic) responses were weaker, implicating atypical suppression in autism.

@@

Experiment 1 was completed by 100 adult participants (32 male, 68 female; mean age 21.9) in early 2015, and first reported by Vilidaite et al. (\protect\hyperlink{ref-Vilidaite2018}{2018}). Here we reanalysed the dataset, and report the results of masking conditions not previously published. Experiment 2 was completed by 100 adult participants (23 male, 74 female, 3 other/not stated; mean age 22.1) in early 2022. Experiment 3 was completed by 10 adults (2 male, 8 female) with a clinical diagnosis of autism, and 10 control participants who were closely matched for age (means of 21.8 and 22, \emph{t} = 0.18, \emph{df} = 18, \emph{p} = 0.86) and exactly matched for gender. Procedures in Experiments 1 and 2 were approved by the ethics committee of the Department of Psychology at the University of York. Procedures for Experiment 3 were approved by the ethics committee of the York Neuroimaging Centre. All participants provided written informed consent, and procedures were consistent with the Declaration of Helsinki.

@@

In Experiment 3, stimuli were presented using a ProPixx DLP projector (VPixx Technologies) running at 120Hz. Stereo presentation was enabled using a circular polariser that was synchronised with the projector refresh, and participants wore passive polarised glasses during the experiment. DLP projectors are perfectly linear, so gamma correction was not required. Data were acquired using a refurbished 248-channel 4D Neuroimaging Magnes 3600 MEG scanner, recording at 1001Hz. Participant head shape was digitised using a Polhemus Fastrak device, and head position was recorded at the start and end of each block by passing current through 5 position coils placed at fiducial points on the head. We also obtained structural MRI scans using a 3 Tesla Siemens Magnetom Prisma scanner to aid in source localisation. Two participants were not available for MRI scans, so we used the MNI ICBM152 template brain (\protect\hyperlink{ref-Fonov2011}{Fonov et al., 2011}) for these individuals.

@@

Participants also completed either the short AQ (\protect\hyperlink{ref-Hoekstra2011}{Hoekstra et al., 2011}) in Experiment 1, or the full AQ (\protect\hyperlink{ref-Baron-Cohen2001}{Baron-Cohen et al., 2001}) in Experiments 2 and 3. For comparison across experiments, we rescaled the short AQ to the same range as the full AQ (0-50). In Experiments 2 and 3, the sensory perception quotient (SPQ) questionnaire (\protect\hyperlink{ref-Tavassoli2014}{Tavassoli et al., 2014}) was also completed.

@@

EEG data from Experiments 1 and 2 were first imported into Matlab using components of the EEGlab toolbox (\protect\hyperlink{ref-Delorme2004}{Delorme \& Makeig, 2004}), and converted into a compressed ASCII format. Primary data analysis was then conducted using a bespoke \emph{R} script. In brief, we epoched each trial and extracted the average timecourse across four occipital electrodes (\emph{Oz}, \emph{POz}, \emph{O1} and \emph{O2}), and then calculated the Fourier transform of this average waveform. We excluded trials for which the Mahalanobis distance of the complex Fourier components exceeded 3 (for details see \protect\hyperlink{ref-Baker2021}{Baker, 2021}). This resulted in 0.25\% of trials being excluded for Experiment 1, and 4.51\% of trials for Experiment 2. Next we averaged the waveforms across all remaining trials, and calculated the Fourier transform in a 1-second sliding window to generate timecourses for each participant. We divided the timecourse for the target-only condition by the timecourse for the target + mask condition to produce a suppression ratio. These were then converted to logarithmic (dB) units for averaging, calculation of standard errors, and statistical comparisons. For display purposes we smoothed the timecourses using a cubic spline function, however all statistical comparisons used the unsmoothed data.

Following the suggestion of a reviewer, we conducted an alternative fixed-phase analysis, where the signal in each 1 second epoch was multiplied by a sine wave of appropriate frequency and phase instead of taking the Fourier transform. The results were similar to our main analysis, and can be viewed in the Figures subdirectory of the project code repository (\href{https://github.com/bakerdh/normreweight/tree/main/Figures}{https://github.com/bakerdh/normreweight/tree/main/Figures files with the suffix `FP'}). We also conducted simulations (also available in the code repository) to confirm that our analysis methods were not distorting the estimates of the suppression timecourse. In brief, although the 1 second sliding time window does blur the signals in time, these effects are largely negated by calculating the suppression ratio because the blur cancels out across the numerator and denominator. Overall these simulations give us confidence in the accuracy of our estimates of suppression dynamics.

For Experiment 3, we performed source localisation using a linearly constrained minimum variance (LCMV) beamformer algorithm, implemented in Brainstorm (\protect\hyperlink{ref-Tadel2011}{Tadel et al., 2011}). Structural MRI scans were processed using Freesurfer (\protect\hyperlink{ref-Dale1999}{Dale et al., 1999}) to generate a 3D mesh of the head and brain, and we calculated source weights for each block with reference to a 5-minute empty room recording, usually recorded on the day of the experiment. The matrix of source weights for each block was used in a custom Matlab script to extract signals from V1, identified using the probabilistic maps of Wang et al. (\protect\hyperlink{ref-Wang2015}{2015}). These signals were then imported into R for the main analysis, which was consistent with the EEG analysis described above. The outlier rejection procedure excluded 2.47\% of trials for Experiment 3.

To make comparisons between groups of participants across time, we used a non-parametric cluster correction technique (\protect\hyperlink{ref-Maris2007}{Maris \& Oostenveld, 2007}) based on t-tests. Clusters were identified as temporally adjacent observations that were all statistically significant, and a summed t-value was calculated for each cluster. A null distribution was then generated by randomising group membership and recalculating the summed t-value for the largest cluster, and repeating this procedure 1000 times. Clusters were considered significant if they fell outside of the 95\% confidence limits of the null distribution. We adapted this approach to test for significantly increasing suppression by conducting one-way t-tests between time points separated by 1000ms, and repeating the cluster correction procedure as described above.

@@

We began by reanalysing data from a steady-state visually evoked potential (SSVEP) experiment reported by Vilidaite et al. (\protect\hyperlink{ref-Vilidaite2018}{2018}). Participants viewed arrays of flickering gratings of varying contrasts. In some conditions a single grating orientation was present flickering at 7Hz (the target), whereas in other conditions a high contrast `mask' was added at right angles to the target gratings, and flickering at 5Hz. The left panel of Figure \ref{fig:Pilotdata}a shows contrast response functions with and without the mask - the presence of the mask reduces the 7Hz response to the target (blue squares are below the black circles; significant main effect of mask contrast, F(1,99) = 26.52, \emph{p} \textless{} 0.001). Similarly, the right panel of Figure \ref{fig:Pilotdata}a shows that the 5Hz response to the mask was itself suppressed by the presence of high contrast targets (main effect of target contrast on the mask response, F(2.92,288.63) = 46.77, \emph{p} \textless{} 0.001; note that the data from the masking conditions were not reported by Vilidaite et al. (\protect\hyperlink{ref-Vilidaite2018}{2018})). At both frequencies, responses were localised to the occipital pole (see insets).

@@

{\centering \includegraphics[width=0.4\linewidth]{Figures/VilidaiteData}

@@

\caption{Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking the electrodes over which signals were averaged (Oz, POz, O1 and O2). Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to a baseline condition (grey) where no stimulus was shown at the target frequency. Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Points around y = 1.8 indicate a significantly increasing ratio (for the time window centred at each point). Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE across N=100 participants, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate the conditions used for subsequent analyses.}\label{fig:Pilotdata}
\end{figure}

We then performed a timecourse analysis, in which we analysed each 11-second trial using a sliding 1-second time window. The top panel of Figure \ref{fig:Pilotdata}c shows the response at the target frequency (7Hz) to a single stimulus of 32\% contrast (black), and the response at 7Hz when the 32\% contrast mask is added (blue). For comparison, a baseline timecourse is also shown (grey), which was the target response at 7Hz when a 5Hz mask stimulus was shown (therefore controlling for attention, blinking etc.). Analogous responses are shown at three other frequencies - the mask frequency (5Hz), and the second harmonics of both target and mask frequencies (14Hz, 10Hz), at which strong responses were also found (see spectra in Figure \ref{fig:Pilotdata}b). The reduction in signal strength when a second component is added at a different orientation and frequency illustrates the masking effect. Surprisingly there was sometimes substantial activity before and after the stimulus was presented, as is especially clear in the baseline condition shown by the grey curves in Figure \ref{fig:Pilotdata}c.~We think the most likely explanation for this is broadband noise from participant movement during the breaks between trials. Since it is approximately equal across conditions it appears to cancel out in the suppression ratios (Figure \ref{fig:Pilotdata}d).
